---
title: "Figs"
author: "JBone"
date: "2025-02-03"
# output: word_document
output:
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    fig_width: 12
    fig_height: 6
    number_sections: TRUE
    fig_caption: TRUE
biblio-style: apalike
bibliography: refs.bib
link-citations: yes
always_allow_html: yes
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, echo=FALSE, root.dir = "/Users/jordanbone/Documents/GitHub/IAV-Mach-Learning",fig.align = "center",fig.height=8,fig.width=10)
source("/Users/jordanbone/Documents/GitHub/IAV-Mach-Learning/MegaLibrary.R")
library(TreeTools)

reference_table <- read_csv("/Users/jordanbone/Documents/GitHub/IAV-Mach-Learning/RefTable.csv")
rep_uids <- read.csv("/Users/jordanbone/Documents/GitHub/IAV-Mach-Learning/USED UIDS.txt")
uid_ref <- read.csv("/Users/jordanbone/Documents/GitHub/IAV-Mach-Learning/USED UIDS.txt")
uid_ref <- inner_join(uid_ref,reference_table[,c(1,14,15)]) %>% distinct()
varimps <- read.csv("/Users/jordanbone/Documents/GitHub/IAV-Mach-Learning/Varimp/VarImp Refs.csv")
```

# Introduction

Influenza A viruses frequently cross host-species boundaries and in
doing so, exemplify zoonotic spillover risks. The potential for IAV
cross-species emergence is one of the most keenly felt infectious
disease threats in the world. The proclivity of these viruses to
circulate in multiple vertebrate species, including many of the species
reared agriculturally, emphasises the importance of understanding how
and when host-jumps may occur. In addition, huge amounts of resources
are spent annually to try and predict whether IAV will jump between
species, and which species and strains may be involved in
(re-)emergence. To predict these cross-species transmission events, many
different aspects of influenza virus biology are routinely surveyed:
evolutionary history, ecological contact networks and structural
modelling of host-pathogen protein interactions, to name but a few.
However, exploring these features independently can obfuscate patterns
and/or lead to repetition of often labour-intensive and costly analyses.
The main aim of this study is to interrogate two methods of observing
host-adaptation in influenza A viruses that have previously jumped from
one species into another. We focus on detecting signal(s) of adaptation
to the original host that linger in the virus that circulates in the
secondary host species. Specifically, phylogenetic trait-based analyses
are compared against molecular and structural characteristics of viral
proteins in order to determine putative benefits of a multi-faceted
approach in understanding (and therefore hopefully, eventually
predicting) host jumps; in other words, is the relationship between
protein structures already accounted for by viral phylogenetic
histories?

Our cross-specialisation approach first uses protein compositional
features to train machine learning models to classify the original host
each virus was obtained from. Viral proteins have long been examined in
order to understand and mitigate the infection process: proteins that
mediate viral entry into host cells, those that usurp cellular machinery
or antagonise cellular defences, and proteins that interact directly
with adaptive immune molecules of the colonised host.

Finally, we compared the results of both modelling and phylogenetic
approaches to detect any overlap or, more interestingly, any inferences
that were exclusive to only one method. In examining two ways to detect
signal of adaptation to an original host, we explore where methods
agree, where they disagree and which is the “best” moving forward.

# Methods

## Sequences

### Sequence Collection

Sequences were collected from NCBI Virus Database. This dataset was
filtered to only include samples with: a) complete nucleotide sequences,
missing no more than 10% of the average length for that segment; b)
metadata containing the virus subtype and original host, at least up to
the Class level (*Aves* or *Mammalia*) through preferably detailing all the
way down to the Family or Species. Nucleotide sequences were then
assigned a random identifier (between AAA-000-AAA and ZZZ-999-ZZZ) to
associate with metadata throughout analyses. 14,682 avian and 21,737
mammalian sequences were collated at this stage.

Samples from mammalian hosts were then retained only if the original host was within the
canine, swine, horse, human or New World bats (*Canidae*, *Suidae*, *Equidae*, *Hominidae* and *Phyllostomidae* respectively)
families, as viruses within these hosts were expected to display clear
examples of host-adaptation as well as having a history of cross-species
transmission. Viral sequences of mammalian-origin were then reduced
further by excluding samples where the subtype does not circulate
naturally within closed systems of that host (i.e. removing spillovers
or stuttering chains of transmission, neither of which represent the
kind of host adaptation we sought to detect). Included subtypes for each
family were: canine (H3N2, H3N8), equine (H7N7, H3N8) and human (H1N1,
H1N2, H2N2 and H3N2). Whole-genome sequences from bats were so sparse
that all examples were retained and the breadth of diversity found in
swine-origin viruses meant that all subtypes were included from this
family.

Corresponding protein sequences were then obtained for each
sample, most of which were available from the NCBI Virus Database. The
few genomic sequences without matching proteins were translated locally
using the ‘ape’ package in R [@paradis_2019aa].

### Clustering

Proteins were then aligned using MAFFT [@rozewicki_2019aa] into 20
resulting protein fasta files (two for each of the 10 major IAV
proteins, one containing mammalian-origin viruses and the other
avian-origin). These were then clustered using the linear algorithm
`easy-linclust` within the tool MMSeq2 [@steinegger_2018aa].
Three variables of the clustering algorithm were experimented with in
optimising data retention: the percentage identity threshold required to
delineate sequences from one another
(`–min-seq-id 0.75, 0.85, 0.9, 0.95, 0.99`), coverage of the sequences
within each cluster (`-c 0.5, 0.7, 0.8`) and the way in which the
coverage is calculated (`–cov-mode 0, 1`). Each possible combination was
tested for avian and mammalian viruses [Supp Table 1].

Options were set to `--min-seq-id 0.95`, `-c 0.8` and `--cov-mode 1` in
order to grant a representative number of samples without introducing
redundancy. Ultimately, 6,697 proteins were retained by clustering
leading to sequences from 4568 individual viruses.

```{r, Unique Subtypes}
subset(uid_ref,!is.na(Subtype)) %>% group_by(Subtype) %>% summarise(Num=length(Subtype)) %>%  ggplot(aes(x=reorder(Subtype,-Num),y=Num+0.1,fill=Subtype %in% held_subtypes))+ geom_bar(stat="identity")+ geom_label(aes(x=17,y=250,label="23 subtypes appear\nmore than 25 times,\n(4260 samples)"),fill="#53CCCD",size=5)+
  geom_label(aes(42,28,label="308 other samples are\ngrouped into the\n'Misc. Subtype' group"),fill="#F8766D",size=5)+
  # geom_label(aes(12,790,label="Holdouts = 4260"),fill="#53CCCD",size=4)+
  geom_segment(aes(x=64,xend=59.15,y=12,yend=2.2),arrow = arrow(angle = 30, length = unit(0.2, "cm"), type = "closed"),colour="black")+ 
  geom_segment(aes(x=76,xend=79.85,y=15,yend=1.2),arrow = arrow(angle = 30, length = unit(0.2, "cm"), type = "closed"),colour="black")+
 geom_label(aes(x=70,y=10,label="Two bat-specific\nsubtypes are included in\nthe 'Holdouts' group"),fill="#53CCCD",size=5)+
  theme(legend.position = "none",
        # text=element_text(size=42), #change font size of all text
        axis.title=element_text(size=12), #change font size of axis titles
        plot.title=element_text(size=14), #change font size of plot title
        # legend.text=element_text(size=20), #change font size of legend text
        # legend.title=element_text(size=20),
        axis.text.x = element_text(size=10.5,angle = 90),axis.text.y = element_text(size=12))+scale_y_log10()+
  labs(title = "4568 IAV Whole-Genome Sequences, Sorted by Subtype",x="Subtype",y="Quantity", caption="Sequence selection collapsed rare virus subtypes into a single group (red bars), which were analysed alongside the 23 most\nfrequently appearing subtypes plus two subtypes originating from bats (blue bars). Thus resulting in a total of 26 viral subgroups used in later analyses.")
ggsave("Used Subtypes.pdf",width=33,height=19,units = "cm")

# subset(uid_ref,!is.na(Subtype)) %>% mutate(Holdout=Subtype %in% held_subtypes) %>% group_by(Holdout) %>% summarise(N=length(Subtype))
```

```{r Sequence Diversity, include=FALSE}
# haBifas <- seqinr::read.fasta("Sequences/Avian/ha_cluster_95_c80_cov0_rep_seq.fasta")
# haBi <- data.frame("Base"=1:567,"Entropy"=Bios2cor::entropy(haBifas),row.names = NULL)
# haMafas <- seqinr::read.fasta("Sequences/Mammals/04HA/HA_cluster_95.fasta")
# haMa <- data.frame("Base"=1:566,"Entropy"=Bios2cor::entropy(haMafas),row.names = NULL)
# 
# ggplot(haBi,aes(Base,Entropy))+geom_line(colour="indianred",alpha=0.8)+geom_line(data=haMa,colour="slateblue",alpha=0.8)+
#   geom_label(aes(300,0.58,label="Avian\nH = 93.13"),colour="indianred")+
#   geom_label(aes(400,0.4,label="Mammalian\nH = 90.30"),colour="slateblue")+labs(title="Sitewise Diversity in Avian and Mammalian Haemagglutinin Clusters")
```

## Model Creation

### Protein Features

Proteins from avian-origin and mammalian-origin viruses retained by the
clustering algorithm within MMSeq2 were then aligned into a single fasta
file for each protein. Accessory proteins (PB1-F2 and PA-X) were then
removed from the analyses as both lie within the coding regions of their
respective genomic segments and thus the calculation of any features
from sequence data would already include these regions. Similarly, those
spliced from alternate reading frames (M2 and NEP) were excluded from
any further analyses as they were too short and/or conserved to show
patterns of adaptation.

Six datasets of protein features were obtained from OmegaFeaturesCLI [@chen_2022aa] which extracted matrices of protein compositional and physio-chemical properties from alignments with the
‘get_descriptor’ command. Features, and the associated command, are as
follows: amino acid two-mers (`DPC type 1`, 400 features), Chou’s pseudo-amino acid composition
(`PseAAC`, 23 features), conjoined triad (`CTriad`, 343 features), and finally Composition, Transition & Distribution values (`CTDc`, `CTDd` and `CTDt`, 39, 195 and 39 features respectively). These were then annotated with the subtype of the
origin virus and labelled with the original host.

### Machine Learning Models

Matrices of protein features were then used to create Random Forest
classification models, to distinguish between the host from which
viruses were sampled.

In order to create test data, subtypes that appeared most frequently
(\>20 times) were iteratively held out of model construction. In
addition to these 23 subtypes, both bat-exclusive subtypes (H17N10 and
H18N11) were held as test data. The remaining 70 subtypes comprised
$\frac{308}{4551}$ (\~7%) sequences. Six Random Forest models were
created from each of the feature datasets, for each holdout. These
models were then stacked with the use of the `caretEnsemble` package
[@deane-mayer_2024aa]. Models were weighted in order to accommodate for
disproportionate sizes of host groups by using the inverse of that
group's proportion.

These datasets were then used to train Random Forest models, leading to
2400 models in total: 8 proteins, 6 chemical and compositional features,
25 subtype holdouts and either two-class or multi-class models. Each
models’ training statistics were extracted for preliminary validation.
Models for each protein and subtype were then compiled with the stacking
algorithm by caret.

![Pipeline of machine learning model construction. One model was created
for each combination of variants, which were then collated into stacked
models to inform and train across broader sample
sizes.](Model Assembly.png)

### Model Performance

Models were assessed with common ML model comparison metrics: accuracy, precision, sensitivity & specificity and balanced accuracy (BA). All performance metrics were carried out courtesy of the `MLmetrics` package [@yan_2025aa].

[This measure of performance ($BA=\frac{TPR+TNR}{2}$) so the closer TPR+TNR is to 1, the worse the model. This is because the model is overestimating positive hits and is unable to distinguish reals from fakes – i.e. which is equivalent to the expected value of a random guess in a balanced data set.]{style="color:blue"}

## Predictive Measures

### Phylogenetics

Using sequence data from each of the 8 proteins studied phylogenetic
trees were estimated. Initial model parameters such as evolutionary and
codon-partitioning models were estimated by ModelFinder within IQTree2
[@kalyaanamoorthy_2017aa; @minh_2020aa], and temporality was confirmed
with TempEst [@rambaut_2016aa].

```{r}
data.frame(Protein=c("PB2","PB1","PA","HA","NP","NA","M1","NS1"),"Model"=c("FLU+R4","FLU+R4","FLU+R5","FLU+R8","Q.mammal+R4","FLU+R5","Q.mammal+R3","HIVw+R6")) %>% kbl() %>% kable_styling(full_width = F, position="float_right")
```

Maximum Likelihood trees were estimated in IQTree2 with the substitution models suggested by ModelFinder (Table 1) and bootstrapped 1000 times for validation.

Phylogenetic models were first trialled on Maximum Likelihood trees with
the use of BayesTraits v3.0 [@meade_2022aa]; in the R wrapper "btw"
[@griffin_2018aa]. First, a discrete-trait analysis estimated the origin host of viruses, protein features were then modelled as continuous traits. Trait analyses were first modelled the features of tips and then ancestral trait reconstruction upon internal nodes, both using multi-state reverse-jump MCMC procedures. Transition rates between discrete states (i.e. Host) were examined across each protein tree before the states of ancestral nodes were estimated.

### Ancestral Trait Reconstruction

Specific nodes were selected for ancestral trait reconstruction due to known historical host-switching events. Recognised host switch events include: human viruses that originated from avian and swine reassortments in 1918 (H1N1), 1957 (H2N2) and 1968 (H3N2); emergence of swine viruses into human pandemic strains (H1N1~pdm09~); avian IAV into horses (H7N7 and H3N8) and the jump of equine H3N8 viruses into canine populations.

```{r, eval=FALSE, echo=TRUE}
####### HA Node 1 #######
nexTree <- read.nexus("HA.nexus")
tree_scaled <- scaleTree(nexTree)
feature_meta <- read.table(paste0("IAV/varimp_HA_values.txt"))

ace_cmd <- c("1","2","RevJump exp 10","AddTag TRecNode ASB30836_BHI882BVL_Canidae_2011 AEX30515_NPS021OCG_Equidae_2010.311111 ACH95671_ZVA445GAI_Suidae_2005.955556 AEM60113_KOI765MSE_Equidae_1979 AEM60147_ATB639FJH_Equidae_1971","AddNode RecNode TRecNode")

features_ACE <- bayestraits(feature_meta, tree_scaled, ace_cmd, silent = FALSE)
features_ACE$Log$results %>% select(-c("No.Off.Parmeters","No.Off.Zero","Model.string",starts_with("Root"))) %>% write.csv(paste0("IAV/HA Node 1 Feature Reconstruction.csv"))
```

# Results

## Dataset Curation

```{r}
data.frame(Protein=c("PB2","PB1","PA","HA","NP","NA","M1","NS1"),"Avian"=c(338,303,282,1406,215,185,73,539),"Mammalian"=c(561,255,415,919,244,207,149,606)) %>% kbl(align='c',caption="Selected sequences") %>% kable_styling(full_width = F, position="float_right")
```

In total 36,419 sequences were found eligible for study (14,682 avian
and 21,737 mammalian). Clustering resulting from removing sequences with a 95% sequence identity left us with 6,697 protein sequences overall. To assist with building the model architecture, any virus genome that had at least one representative protein was analysed fully - hence 4,551 viruses were analysed, despite the high sequence homology between many of the proteins except for the selected representative protein. The number of representative proteins called by the clustering algorithm is shown in Table 2 and full lists of the associated accession codes are available in Supplementary 1.

## Protein Models

Confusion matrices were made on the known origin host and the host predicted by the machine learning models. Matrices were used to assess model predictive strength. Incorrectly labelled viruses were indicative of signal of possible adaptation to the host from which a virus originated.

```{r, Model stat table}
data.frame(Protein=c("PB2","PB1","PA","HA","NP","NA","M1","NS1"),"F1 micro"=round(c(0.8431141,0.8074284,0.8750299,0.7699953,0.8840125,0.7349682,0.8632233,0.909803),3),"F1 macro"=round(c(0.5176765,0.3840304,0.5749008,0.2947088,0.495701,0.332125,0.5168716,0.51886),3),"AUC"=percent(c(0.57,0.527,0.575,0.558,0.556,0.527,0.559,0.565),accuracy=0.01)) %>% kbl(align='c',caption="Model Validation") %>% kable_styling(full_width = F, position="float_right")
```

Counter-intuitively, a lower model accuracy implied greater evidence of adaptation to the host species; the mislabelling of proteins is indicative of host-specific adaptations to non-native hosts. For example, should a human-origin virus be incorrectly labelled as swine-origin by the model, this supports the hypothesis that the protein retains some signal of adaptation to the original swine hosts. Overall, confusion matrices were evaluated based on accuracy and F1 statistics (shown in Table 3).

```{r, ConfMatrix Performance}
master_conf <- readRDS("/Users/jordanbone/Documents/GitHub/IAV-Mach-Learning/Comp/ConfMatrix Multiclass Master.rds")

master_conf$table %>% as.data.frame() %>% ggplot(aes(Prediction,Reference,fill=Freq/sum(master_conf$table)))+ geom_tile(colour="grey10")+
  geom_text(aes(label=ifelse(Freq<=0.01," ",percent(Freq/sum(master_conf$table),accuracy = 0.01))))+
  scale_fill_gradient(low="white",high="forestgreen",na.value = "grey50")+
  guides(fill=guide_legend(title="Freq %"))+labs(caption="Total confidence matrix of all 2400 models, averaged across each of the 8 proteins. Values greater than 0.01% are labelled to emphasise mismatches.")
#ggsave("Mislabelling.png",width=20,height=20,units = "cm")
```
[Main outliers (>1%) are pig-human-avian though maybe just caused by the relative proportion of these. The RF models do show good estimation of true-positives, especially for the most frequently appearing samples. Interestingly though, the biggest mistake that the model made was of misclassifying human-origin sequences as pig-origin. In fact, more of the human sequences were incorrectly classified as swine (8.13%) than correctly identified (7.23%).]{style="color:blue"}

```{r}
classes <- master_conf$byClass %>% as.data.frame() %>% rownames_to_column()
classes$Host <- c("Avian","Dog","Horse","Human","Bat","Swine")
m1 <- ggplot(classes,aes(Sensitivity,Specificity,colour=Host))+geom_point(size=3)+ylim(0,1)+labs(title = "Performance")
m2 <- ggplot(classes,aes(Host,Precision,colour=Host))+geom_point(size=3)+ylim(0,1)+theme(legend.position="none")+labs(title = "Precision")
m3 <- ggplot(classes,aes(Host,`Balanced Accuracy`,colour=Host))+ylim(0,1)+geom_hline(yintercept = 0.5,linetype="longdash",colour="grey")+geom_point(size=3)+theme(legend.position="none")+labs(title = "Error Accuracy")
m4 <- ggplot(classes,aes(`Pos Pred Value`,`Neg Pred Value`,colour=Host))+geom_point(size=3)+ylim(0,1)+labs(title = "Predictive Values",x="Positive Pred. Value",y="Negative Pred. Value")

ggarrange(m1,m4,m3,m2,common.legend = T, legend="bottom", labels="AUTO") %>% annotate_figure(bottom="A) Performance, B) Model accuracy, C) The Balanced Accuracy (BA) weighs errors that the model makes, and\nwhether they are biased towards a certain type of error (false-positives or false-negatives) and D) Predictive precision")
ggsave("Model Diag.png",width=20,height=20,units = "cm")
```

Analysing the results of random forest classification models based on the protein characteristics alone, predicted host class varied heavily depending on the host. First, all models have high (>75%) specificity but sensitivity varies wildly. This isn’t helped by the sparsity of canine, equine and bat samples in the dataset.

Again, high negative predictive and varying positive predictive values are likely influenced by the relatively small number of sequences from viruses originating from bat, dog and horse hosts. Weirdly, proteins from bat viruses seem to perform better than those from canine and equine hosts; maybe the sparsity of proteins from dog and horse viruses throw things off more than the bats which are highly divergent to other IAV. However, both of these first two measures imply that host signal is evident in the protein chemical features but extrapolating this to actual cross-species jumps is less reliable.

In model performance, dog-origin viruses are the only time that the models are basically guessing randomly, reflected in the Balanced Accuracy.

[Overall though, the models perform pretty well given how complex this is and the fact that it is just based on 1039 protein features alone]{style="color:blue"}

### Protein Features

By running ML models and iteratively removing each protein feature, the model performance statistics can be used to infer which features cause performance to drop the greatest. Hence, taking out the features and re-running the model allows us assess the impact that that feature had on model performance. The drop in AUC is used to assess the importance of each feature within each ML model.

```{r,eval=FALSE}
  overall_2c <- c()
  byClass_2c <- c()
  overall_Mc <- c()
  byClass_Mc <- c()
for (PRT in c("PB2","PB1","PA","HA","NP","NA","M1","NS1")){
  cm2c <- readRDS(paste0("Comp/",PRT," 2class ConfMatrix.rds"))
  cmMc <- readRDS(paste0("Comp/",PRT," Mclass ConfMatrix.rds"))
  
  # cm2c$table
  ovr2 <- data.frame(PRT,t(cm2c$overall))
  byc2 <- data.frame(PRT,Val=cm2c$byClass) %>% rownames_to_column()
  
  # cmMc$table
  ovrM <- data.frame(PRT,t(cmMc$overall))
  bycM <- data.frame(PRT,cmMc$byClass) %>% rownames_to_column()
  
  overall_2c <- bind_rows(overall_2c,ovr2)
  byClass_2c <- bind_rows(byClass_2c,byc2)
  overall_Mc <- bind_rows(overall_Mc,ovrM)
  byClass_Mc <- bind_rows(byClass_Mc,bycM)
}

overall_2c
byClass_2c <- byClass_2c %>% pivot_wider(names_from = rowname,values_from = Val)
overall_Mc
byClass_Mc$rowname <- byClass_Mc$rowname %>% str_split_i(":",2) %>% str_trim()
```

```{r, eval=FALSE}
mc_byc <- read.csv("../Comp/Multiclass breakdown.csv")
ggplot(mc_byc,aes(Sensitivity,Specificity,colour=Protein))+geom_point(alpha=0.8)+facet_wrap(~Host)+theme(legend.position = "bottom")
#ggsave("HostDiag1.png",width=20,height=20,units = "cm")
ggplot(mc_byc,aes(Pos.Pred.Value,Neg.Pred.Value,colour=Protein))+geom_point()+facet_wrap(~Host)+theme(legend.position = "bottom")
#ggsave("HostDiag2.png",width=20,height=20,units = "cm")
```

```{r, eval=FALSE}
ggplot(mc_byc,aes(Host,Precision,colour=Protein))+geom_jitter(height=0,width=0.2,size=2)+labs(title="ML Protein Model Performance")
#ggsave("Model Precision.png",width=20,height=20,units = "cm")

ggplot(mc_byc,aes(Host,F1,colour=Protein))+geom_jitter(height=0, width=0.2,size=2)
#ggsave("Model F1.png",width=20,height=20,units = "cm")

ggplot(mc_byc,aes(Host,Balanced.Accuracy,colour=Protein))+geom_jitter(height=0, width=0.2,size=2)
#ggsave("Model BalAcc.png",width=20,height=20,units = "cm")

ggplot(mc_byc,aes(Protein,Host,fill=F1))+geom_tile()+geom_text(aes(label=percent(F1,accuracy=0.01)))+
  scale_fill_gradient(low="grey90",high="forestgreen",na.value = "white")+ theme(legend.position="none")+ labs(title = "F1 Values of Multiclass Models")

ggplot(mc_byc,aes(Protein,Host,fill=Balanced.Accuracy))+geom_tile()+scale_fill_gradient(low="grey90",high="forestgreen")+geom_text(aes(label=percent(F1,accuracy=0.01)))+
  scale_fill_gradient(low="grey90",high="forestgreen",na.value = "white")+ theme(legend.position="none")+ labs(title = "Balanced Accuracy of Multiclass Models")
```

## Mismatches

```{r, eval=FALSE}
for (PRT in c("PB2","PB1","PA","HA","NP","NA","M1","NS1")){
  tre <- read.nexus(paste0("Phylogenies/iqtree/trees/NEXUS/",PRT,".nexus"))
  pre <- read.csv(paste0("Comp/",PRT,"_predictions.csv"))
  
  if(PRT=="HA"){
    labs <- data.frame(tre$tip.label,V1=tre$tip.label %>% str_split_i("_",2))
  }
  else{labs <- data.frame(tre$tip.label,V1=tre$tip.label %>% str_split_i("_",1))}
  left_join(labs,pre,by="V1",keep = F) %>% mutate(Mismatch=V4==Prediction) %>%
    write.csv(paste0("Phylogenies/iqtree/Annotations/",PRT,".txt"), row.names = F,quote = F)
}
```

```{r}
mismatches <- data.frame()
for (PRT in c("PB2","PB1","PA","HA","NP","NA","M1","NS1")){
  msmtch <- read.csv(paste0("../Phylogenies/iqtree/Annotations/",PRT,".txt"),sep="\t")
  mismatches <- rbind(mismatches,msmtch)
}
mismatches$PRT <- mismatches$PRT %>% str_replace_all("PB2","01PB2") %>% str_replace_all("PB1","02PB1") %>% str_replace_all("PA","03PA") %>% str_replace_all("HA","04HA") %>% str_replace_all("NP","05NP") %>% str_replace_all("M1","07M1") %>% str_replace_all("NS1","08NS1") %>% replace_na("06NA")

mismatches %>% group_by(PRT,Matched) %>% summarise(N=length(Matched)) %>% subset(!is.na(Matched)) %>% pivot_wider(names_from = Matched, values_from = N) %>% 
  mutate(TP=percent(`TRUE`/(`TRUE`+`FALSE`)),FP=percent(`FALSE`/(`TRUE`+`FALSE`))) %>% 
  select(PRT,`TRUE`,TP,`FALSE`,FP) %>% kbl(col.names = c("Protein","raw","%","raw","%"),align='c',caption="Mislabelled") %>% kable_styling(full_width = F,position="float_left") %>% add_header_above(c(" ", "Matched"=2, "Mismatched"=2))
```

[When did the model predict things incorrectly? Was there any pattern to the mistakes? Not really...]{style="color:blue"}

## Feature Importance

[This is the variable importance stuff, explain better how it was done
and what the AUC loss means]{style="color:blue"}

By running ML models and iteratively removing each protein feature, the model performance statistics can be used to infer which features cause performance to drop the greatest. Hence, taking out the features and re-running the model allows us assess the impact that feature had on model performance.

```{r}
var_imps <- read.csv("../Varimp/VarImps.csv")

# ggplot(subset(var_imps,AUC_loss>0),aes(AUC_loss,colour = Protein))+ geom_density()+ ylim(0,500)
# ggplot(subset(var_imps,AUC_loss>0.001),aes(AUC_loss,colour = Protein))+ geom_density()
##ggsave("Figures & Presentables/VarImp1.pdf",width=33,height=19,units = "cm")

ggplot(subset(var_imps,AUC_loss>0),aes(AUC_loss,colour = Protein))+ geom_density()+facet_wrap(~str_split_i(var,"_",1),scales="free")
#ggsave("AUC Loss.png",width=20,height=20,units = "cm")
```

```{r, eval=FALSE}
ggplot(subset(var_imps,var %>% startsWith("PAAC_Xc2.lambda2")),aes(Protein,AUC_loss,colour = Protein,size=Protein=="01PB2"|Protein=="03PA"))+ geom_point()+geom_hline(yintercept = 0,colour="grey20",linetype="longdash")+ theme(legend.position = "none")+labs(title="PAAC_Xc2.lambda2")
##ggsave("Figures & Presentables/VarImp1.pdf",width=33,height=19,units = "cm")

ggplot(subset(var_imps,var %>% startsWith("PAAC_Xc1.Q")),aes(Protein,AUC_loss,colour = Protein,size=Protein=="05NP"|Protein=="03PA"))+ geom_point()+geom_hline(yintercept = 0,colour="grey20",linetype="longdash")+ theme(legend.position = "none")+labs(title="PAAC_Xc1.Q")
##ggsave("Figures & Presentables/VarImp2.pdf",width=33,height=19,units = "cm")

ggplot(subset(var_imps,var %>% startsWith("PAAC_Xc1.E")),aes(Protein,AUC_loss,colour = Protein,size=Protein=="05NP"|Protein=="06NA"))+ geom_point()+geom_hline(yintercept = 0,colour="grey20",linetype="longdash")+ theme(legend.position = "none")+labs(title="PAAC_Xc1.E")
##ggsave("Figures & Presentables/VarImp3.pdf",width=33,height=19,units = "cm")

ggplot(subset(var_imps,var %>% startsWith("CTDD_hydrophobicity_ENGD860101.1.residue25")), aes(Protein,AUC_loss,colour = Protein,size=Protein=="05NP"|Protein=="07M1"))+ geom_point()+geom_hline(yintercept = 0,colour="grey20",linetype="longdash")+ theme(legend.position = "none")+labs(title="CTDD_hydrophobicity_ENGD860101.1.residue25")
##ggsave("Figures & Presentables/VarImp4.pdf",width=33,height=19,units = "cm")

ggplot(subset(var_imps,var %>% startsWith("CTDC_normwaalsvolume.G2")),aes(Protein,AUC_loss,colour = Protein,size=Protein=="06NA"|Protein=="07M1"))+ geom_point()+geom_hline(yintercept = 0,colour="grey20",linetype="longdash")+ theme(legend.position = "none")+labs(title="CTDC_normwaalsvolume.G2")
##ggsave("Figures & Presentables/VarImp5.pdf",width=33,height=19,units = "cm")
```

## Phylogenetics

Alignments of each of the eight proteins were used to estimate trees
using IQTree2.

[Could also stick some quick diversity measures in for each alignment, to
show what we're working with?]{style="color:blue"}

[How correlated are protein features, length and
diversity?]{style="color:blue"}

```{r, Diversity, eval=FALSE}
for (PRT in c("PB2","PB1","PA","HA","NA","M1","NS1")){
  tree_temp <- read.alignment(paste0("../Phylogenies/Alignments/",PRT,".nexus"))
  tip <- as.phylo(tree_temp)$tip.label
}

pb2 <- read.alignment("../Phylogenies/Alignments/PB2_alignment.fasta",format = 'fasta')
```

### Protein Trees

[Is it worth adding bootstrap values? Or describing the tree (num of tips/nodes) just to clarify that some sequences were excluded from tree estimation?]{style="color:blue"}

### Tree with Protein Features

```{r, eval=FALSE}
for (PRT in c("PB2","PB1","PA","HA","NA","M1","NS1")){
  tree_temp <- read.nexus(paste0("../Phylogenies/iqtree/trees/NEXUS/",PRT,".nexus"))
  tip <- as.phylo(tree_temp)$tip.label
  tA <- data.frame(node = nodeid(tree_temp, tip[grep("Avian", tip)]), host = "Avian")
  tB <- data.frame(node = nodeid(tree_temp, tip[grep("Phyllostomidae", tip)]), host = "Bat")
  tD <- data.frame(node = nodeid(tree_temp, tip[grep("Canidae", tip)]), host = "Dog")
  tE <- data.frame(node = nodeid(tree_temp, tip[grep("Equidae", tip)]), host = "Horse")
  tH <- data.frame(node = nodeid(tree_temp, tip[grep("Hominidae", tip)]), host = "Human")
  tP <- data.frame(node = nodeid(tree_temp, tip[grep("Suidae", tip)]), host = "Pig")
  d <- rbind(tA,rbind(tB,rbind(tD,rbind(tE,rbind(tH,tP)))))
  tree <- full_join(tree_temp, d, by = 'node')
  assign(paste0(PRT,"_tree"),tree)
  saveRDS(tree,file = paste0("../Phylogenies/",PRT,"_tree.rds"))
}
```

```{r, eval=FALSE}
readRDS("../Phylogenies/PB2_tree.rds") %>% ggtree(aes(colour=host))+ 
  # geom_nodelab(aes(label = node), hjust = -0.5)+ geom_tiplab(hjust = -.1)+ geom_tippoint()+
  theme(legend.position = c(.05, .75))+labs(title = "PB2 Tree")
#ggsave("PB2 Tree.png",width=10,height=10,units = "cm")

readRDS("../Phylogenies/PB1_tree.rds") %>% ggtree(aes(colour=host))+ 
  # geom_nodelab(aes(label = node), hjust = -0.5)+ geom_tiplab(hjust = -.1)+ geom_tippoint()+
  theme(legend.position = c(.05, .75))+labs(title = "PB1 Tree")
#ggsave("PB1 Tree.png",width=10,height=10,units = "cm")

readRDS("../Phylogenies/PA_tree.rds") %>% ggtree(aes(colour=host))+ 
  # geom_nodelab(aes(label = node), hjust = -0.5)+ geom_tiplab(hjust = -.1)+ geom_tippoint()+
  theme(legend.position = c(.05, .75))+labs(title = "PA Tree")
#ggsave("PA Tree.png",width=10,height=10,units = "cm")

readRDS("../Phylogenies/HA_tree.rds") %>% ggtree(aes(colour=host))+ 
  # geom_nodelab(aes(label = node), hjust = -0.5)+ geom_tiplab(hjust = -.1)+ geom_tippoint()+
  theme(legend.position = c(.05, .75))+labs(title = "HA Tree")
#ggsave("HA Tree.png",width=10,height=10,units = "cm")

readRDS("../Phylogenies/NP_tree.rds") %>% ggtree(aes(colour=host))+ 
  # geom_nodelab(aes(label = node), hjust = -0.5)+ geom_tiplab(hjust = -.1)+ geom_tippoint()+
  theme(legend.position = c(.05, .75))+labs(title = "NP Tree")
#ggsave("NP Tree.png",width=10,height=10,units = "cm")

readRDS("../Phylogenies/NA_tree.rds") %>% ggtree(aes(colour=host))+ 
  # geom_nodelab(aes(label = node), hjust = -0.5)+ geom_tiplab(hjust = -.1)+ geom_tippoint()+
  theme(legend.position = c(.05, .75))+labs(title = "NA Tree")
#ggsave("NA Tree.png",width=10,height=10,units = "cm")

readRDS("../Phylogenies/M1_tree.rds") %>% ggtree(aes(colour=host))+
  # geom_nodelab(aes(label = node), hjust = -0.5)+ geom_tiplab(hjust = -.1)+ geom_tippoint()+
  theme(legend.position = c(.05, .75))+labs(title = "M1 Tree")
#ggsave("M1 Tree.png",width=10,height=10,units = "cm")

readRDS("../Phylogenies/NS1_tree.rds") %>% ggtree(aes(colour=host))+
  # geom_nodelab(aes(label = node), hjust = -0.5)+ geom_tiplab(hjust = -.1)+ geom_tippoint()+
  theme(legend.position = c(.05, .75))+labs(title = "NS1 Tree")
#ggsave("NS1 Tree.png",width=10,height=10,units = "cm")
```

```{r}
library(nodiv)
tree <- readRDS("../Phylogenies/NS1_tree.rds")
# ggtree(tree,aes(colour=host))+
  # geom_balance(node = 1621)+
  # geom_hilight(node = 1674)+
#   geom_hilight(node = 1621)+
#   theme(legend.position = c(.05, .75))+labs(title = "NS1 Tree")

# tree_subset(tree,1674,levels_back=0) %>% ggtree(aes(colour=host))+ geom_nodelab(aes(label=node),hjust = -0.5)

# basal_node(as.phylo(tree))
# nodenumbers(as.phylo(tree))
# nodes(as.phylo(tree), all = FALSE)
# Descendants(1621, as.phylo(tree))
# Parent("ATB639FJH_AEM60152_Equidae_1971", as.phylo(tree)) #NS1 - 1674
# MostRecentAncestor(tips=c("OWQ498HSV_AHL21355_Suidae_2011.894444","FTK531MEX_ARV89158_Suidae_2014.555556"),as.phylo(tree)) 
# Parent(1672, as.phylo(tree))
```

```{r, eval=FALSE}
HA_tree <- readRDS("../Phylogenies/HA_tree.rds")
ggtree(HA_tree,aes(colour=host))+ geom_nodelab(aes(label=node),hjust = -0.5)+ theme(legend.position = c(.05, .75))+labs(title = "HA Tree")

tree_subset(HA_tree,"AEM60113_KOI765MSE_Equidae_1979",levels_back=2) %>% ggtree(aes(colour=host))+ geom_nodelab(aes(label=node),hjust = -0.5)
```

```{r, fig.height=10}
ha <- readRDS("../Phylogenies/HA_tree.rds")
tip.label(ha) <- tip.label(ha) %>% str_replace_all("'","")
td <- read.table("../Phylogenies/iqtree/Annotations/varimp_HA_values.txt")
ha <- full_join(ha,data.frame(label=td$V1,subtype=td$V13), by = 'label')

p <- ggtree(ha,aes(colour=host))+ geom_tippoint()+ theme(legend.position = c(.05,.75))+ labs(title = "HA Tree") + geom_facet(panel="CTDT charge.Tr1221", data=td, geom=geom_point, mapping=aes(x=V3))

# p
# ggsave("Figures & Presentables/HA Tree with Feature.png",width=12,height=14,units = "cm")

# ggplot(td,aes(V1,V3))+geom_point()+ scale_y_continuous(c(0.1,0.2))

# ggtree(ha,aes(colour=subtype),branch.length = "none")+ geom_tiplab(aes(lable=subtype))+geom_nodelab(aes(label=node),hjust = -0.25)+ theme(legend.position = c(.05,.75))+ labs(title = "HA Tree")
# #ggsave("Phylogenies/iqtree/Annotations/HA_values.pdf",height=42)

subtype_block <- data.frame(node=c(2579,2569,3446,2008,3770,3186,3988,3429), offset=0.003,offset.text=-0.001,extend=0.05, label=c("1","2","3","5","7","9","10","11"))

p + geom_cladelab(data=subtype_block, mapping = aes(node=node, label=label, offset=offset, offset.text=offset.text, extend=extend),barsize = 1,fontface = 3,align = TRUE)
# ggsave("Figures & Presentables/HA Tree with Feature.png",width=12,height=14,units = "cm")
```

[Is there correlation between charge and branch length/placement? Would
need a nice branch-branch measurement like Faith's PD]{style="color:blue"}

```{r, fig.height=10, eval=FALSE}
ggtree(ha,aes(colour=host))+ geom_tippoint()+ theme(legend.position = c(.05,.75))+ labs(title = "HA Tree") + geom_facet(panel="CTDT secondarystruct.Tr1221", data=td, geom=geom_point, mapping=aes(x=V4))+ geom_facet(panel="CTDT secondarystruct.Tr1221", data=td, geom=geom_path, alpha=0.6, mapping=aes(x=V4))

ggtree(ha,aes(colour=host))+ geom_tippoint()+ theme(legend.position = c(.05,.75))+ labs(title = "HA Tree") + geom_facet(panel="CTDT solventaccess.Tr1331", data=td, geom=geom_point, mapping=aes(x=V5))+ geom_facet(panel="CTDT solventaccess.Tr1331", data=td, geom=geom_path, alpha=0.6, mapping=aes(x=V5))

ggtree(ha,aes(colour=host))+ geom_tippoint()+ theme(legend.position = c(.05,.75))+ labs(title = "HA Tree") + geom_facet(panel="CTDT hydrophobicity_CASG920101.Tr1331", data=td, geom=geom_point, mapping=aes(x=V6))+ geom_facet(panel="CTDT hydrophobicity_CASG920101.Tr1331", data=td, geom=geom_path, alpha=0.6, mapping=aes(x=V6))
```

```{r, eval=FALSE}
m1 <- readRDS("../Phylogenies/M1_tree.rds")
tip.label(m1) <- tip.label(m1) %>% str_replace_all("'","")
td <- read.table("Phylogenies/iqtree/Annotations/varimp_M1_values.txt")

ggtree(m1,aes(colour=host))+ geom_tiplab(hjust = -.1)+ theme(legend.position = c(.05,.75))+ labs(title = "M1 Tree") + geom_facet(panel="V1", data=td, geom=geom_point, mapping=aes(x=V3))+ geom_facet(panel="CTDD hydrophobicity_ENGD860101.1.residue25", data=td, geom=geom_path, alpha=0.6, mapping=aes(x=V3))
```

### Trait Analyses

Transition rates between hosts, based on entire trees when the ONLY
feature input to the phylogenetic model is host. i.e. given the
distribution of hosts throughout the tree, what is the likelihood that
any given branch will transition from host~1~ to host~2~?

```{r}
traits <- data.frame()
for (PRT in c("PB2","PB1","PA","HA","NP","NA","M1","NS1")){
  temp <- read.csv(paste0("../Phylogenies/iqtree/Trait Analyses/",PRT," MultiState Trait Analysis.csv"))
  trait_temp <- temp %>% pivot_longer(cols=4:length(temp),names_to = "Transition",values_to = "Rate") %>% mutate(Protein=PRT,Recip=Transition %>% str_replace_all("q","") %>% str_sub(1,1),Donor=Transition %>% str_replace_all("q","") %>% str_sub(2,2))
  traits <- rbind(traits,trait_temp)
}
traits <- traits %>% select(-Tree.No)
traits$Protein <- traits$Protein %>% str_replace_all("PB2","01PB2") %>% str_replace_all("PB1","02PB1") %>% str_replace_all("PA","03PA") %>% str_replace_all("HA","04HA") %>% str_replace_all("NP","05NP") %>% str_replace_all("NA","06NA") %>% str_replace_all("M1","07M1") %>% str_replace_all("NS1","08NS1")
traits$Recip <- traits$Recip %>% str_replace_all("A","Avian") %>% str_replace_all("B","Bat") %>% str_replace_all("D","Dog") %>% str_replace_all("H","Human") %>% str_replace_all("P","Pig") %>% str_replace_all("E","Horse")
traits$Donor <- traits$Donor %>% str_replace_all("A","Avian") %>% str_replace_all("B","Bat") %>% str_replace_all("D","Dog") %>% str_replace_all("H","Human") %>% str_replace_all("P","Pig") %>% str_replace_all("E","Horse")

trait_labs <- traits %>% group_by(Protein,Donor,Recip) %>% summarise(mRate=mean(Rate),sdRate=sd(Rate))
```

```{r, eval=FALSE}
# ggplot(trait_labs,aes(Donor,Recip,fill=mRate))+geom_tile(colour="white")+ scale_fill_gradient(low="grey90",high="forestgreen")+ 
#   geom_text(aes(label=paste0(round(mRate,2),"\n±",round(sdRate,2))),size=2)+
#   facet_wrap(~Protein,ncol=4)+theme_bw()+theme(legend.position = "none")+labs(title = "Discrete Trait Transition Rates of IAV Proteins",subtitle = "Modelled by rjMCMC Multistate Estimation in BayesTraits")
##ggsave("Host Transitions.pdf",width=33,height=19,units = "cm")

ggplot(subset(trait_labs,Protein %in% c("01PB2","02PB1","03PA","04HA")), aes(Donor,Recip,fill=mRate))+ geom_tile(colour="white")+ scale_fill_gradient(low="grey90",high="forestgreen")+ geom_text(aes(label=ifelse(mRate<1,"",paste0(round(mRate,2),"\n±",round(sdRate,2)))),size=2)+facet_wrap(~Protein,ncol=2)+theme_bw()+theme(legend.position = "none")+labs(title = "Discrete Trait Transition Rates of IAV Proteins",subtitle = "Modelled by rjMCMC Multistate Estimation in BayesTraits")
#ggsave("Host Transitions 1.png",width=12,height=12,units = "cm")

ggplot(subset(trait_labs,Protein %in% c("05NP","06NA","07M1","08NS1")), aes(Donor,Recip,fill=mRate))+ geom_tile(colour="white")+ scale_fill_gradient(low="grey90",high="forestgreen")+  geom_text(aes(label=ifelse(mRate<1,"",paste0(round(mRate,2),"\n±",round(sdRate,2)))),size=2)+facet_wrap(~Protein,ncol=2)+theme_bw()+theme(legend.position = "none")+labs(title = "Discrete Trait Transition Rates of IAV Proteins",subtitle = "Modelled by rjMCMC Multistate Estimation in BayesTraits")
#ggsave("Host Transitions 2.png",width=12,height=12,units = "cm")
# ggplot(trait_labs,aes(paste0(Donor,Recip),y=mRate,ymin=ifelse(mRate-sdRate>0,mRate-sdRate,0),ymax=mRate+sdRate,colour=Donor))+geom_pointrange(size=0.4)+facet_wrap(~Protein,ncol=4)+theme_bw()+theme(legend.position = "bottom")

ggplot(trait_labs,aes(Donor,Recip,fill=mRate))+ geom_tile(colour="white")+ scale_fill_gradient(low="grey90",high="forestgreen")+ geom_text(aes(label=ifelse(mRate<1,"",paste0(round(mRate,2),"\n(±",round(sdRate,2),")"))),size=2)+facet_wrap(~Protein,ncol=4)+theme_bw()+
  theme(legend.position = "none",axis.text.x = element_text(size=8),axis.text.y = element_text(size=8))+
  labs(title = "Discrete Trait Transition Rates of IAV Proteins",subtitle = "Modelled by rjMCMC Multistate Estimation in BayesTraits",y="Recipient Host",x="Origin Host")
ggsave("Host Transition Rate.pdf",width=26,height=12,units = "cm")
```


```{r}
ggplot(trait_labs,aes(Donor,Recip,fill=mRate))+ geom_tile(colour="white")+ scale_fill_gradient(low="grey90",high="limegreen")+ geom_text(aes(label=ifelse(mRate<1,"",round(mRate,2))),size=3)+facet_wrap(~Protein,ncol=4)+theme_bw()+
  theme(legend.position = "none",axis.text.x = element_text(size=8),axis.text.y = element_text(size=8))+
  labs(title = "Discrete Trait Transition Rates of IAV Proteins",subtitle = "Modelled by rjMCMC Multistate Estimation in BayesTraits",y="Recipient Host",x="Origin Host")
ggsave("Host Transition Rate v2.pdf",width=26,height=12,units = "cm")
```

[Interesting how rarely avian is estimated as the donor, considering reality]{style="color:blue"}

[Wonder if the pig -\> human values would drop if pdm_2009 were held as a separate group of sequences]{style="color:blue"}

[Nice horse-dog rates given how closely related they are]{style="color:blue"}

[Also interesting when matrices are asymmetrical, indicates a direction of transfer from the phylogenetic signal]{style="color:blue"}

### Ancestral Reconstruction

[Describe subtree selection:]{style="color:blue"}

  - node number

  - number of tips

  - why were these subtrees selected

Bayestraits ACE output gives the model iteration, the current likelihood of the chain, tree number, the features used in model training (10 protein features, *0 - 9*), the labelled traits (six host types, *A, B, D, E, H, P*), actual transition rates (*q~0,A~ - q~9,P~*) and finally the traits estimated from the preceding values.

```{r, eval=FALSE}
transition_rates <- data.frame()
RecNodes <- data.frame()
for (PRT in c("PB2","PB1","PA","HA","NA","M1","NS1")){
  raw_ace <- read.csv(paste0("../Phylogenies/iqtree/ACE/",PRT," Trait Reconstruction.csv"))
  rates <- raw_ace %>% select(Iteration,Lh,starts_with("q")) %>% pivot_longer(cols=3:32,names_to = "Transition",values_to = "Rate") %>% mutate(Protein=PRT,Donor=Transition %>% str_replace_all("q","") %>% str_sub(1,1),Recip=Transition %>% str_replace_all("q","") %>% str_sub(2,2))
  transition_rates <- rbind(transition_rates,rates)
  
  recons <- raw_ace %>% select(Iteration,Lh,starts_with("RecNode")) %>% pivot_longer(cols=3:8,names_to = "RecTrait",values_to = "RecNode") %>% mutate(Protein=PRT,RecTrait=RecTrait %>% str_split_i("\\.",3))
  RecNodes <- rbind(RecNodes,recons)
}
# RecNodes$RecTrait <- ifelse(RecNodes$RecTrait="A","Avian","B" "D" "E" "H" "P")

trat <- transition_rates %>% group_by(Protein,Transition=str_replace_all(Transition,"q","")) %>% summarise(mRate=mean(Rate),sdRate=sd(Rate))
# trat <- transition_rates %>% group_by(Protein,Transition) %>% summarise(Rate=quantile(Rate))
# trat$Stat <- rep(c("min","q25","med","q75","max"),210)
```

```{r}
ace_nodes <- list.files(path = "../Phylogenies/iqtree/ACE",pattern = "Node",full.names = T)

for(i in 1:6){
  ace_data <- read.csv(ace_nodes[i])
  prot <- ace_nodes[i] %>% str_split_i("\\/",4) %>% str_split_i(" ",1)
  node <- ace_nodes[i] %>% str_split_i("\\/",4) %>% str_split_i(" ",3)
    x <- ace_data %>% select(starts_with("RecNode")&(ends_with("P.A.")|ends_with("P.B.")|ends_with("P.D.")|ends_with("P.E.")|ends_with("P.H.")|ends_with("P.P."))) %>% pivot_longer(cols=1:66,names_to = "Rec",values_to = "Value") %>% mutate("Feat"=str_split_i(Rec,"\\.\\.\\.",2) %>% str_split_i("\\.",2),"Host"=str_split_i(Rec,"\\.\\.\\.",3) %>% str_split_i("\\.",3)) %>% group_by(Host,Feat=as.integer(Feat)) %>% summarise(mV=mean(Value),sV=sd(Value),minV=quantile(Value)[1],q25V=quantile(Value)[2],medV=quantile(Value)[3],q75V=quantile(Value)[4],maxV=quantile(Value)[5]) %>% mutate(Protein=prot,Node=node)
    x <- left_join(x,varimps,by=c("Protein","Feat"))
    x$Feat <- str_replace_all(x$Feat,"10","host_ft")
    x$Feature <- x$Feature %>% str_replace_all("_","\n")
    assign(paste0(prot,"_",node),x)
}
```

```{r,fig.height=6}
ace_example <- read.csv("../Phylogenies/iqtree/ACE/HA Node 1 Feature Reconstruction.csv")
head(ace_example[,1:12])
ggplot(ace_example,aes(Iteration,Lh))+geom_smooth(method="lm")+geom_point()+
  labs(y="Chain Likelihood",caption="This example of the chain likelihoods (taken from the BayesTraits output of node 1 from the HA tree) represent the performance of the trait analysis in this subtree.")
```

### Trees & Subtrees

```{r NS1 subtrees, fig.height=10}
NS1_tree <- readRDS("../Phylogenies/NS1_tree.rds")
  
NS1_full <- ggtree(NS1_tree,aes(colour=host))+ 
  # geom_nodelab(aes(label=node),hjust = -0.25)+
  geom_hilight(node = 1673)+geom_hilight(node = 1256)+
  geom_cladelab(node=1673,label="1",angle=270,offset=0.2)+
  geom_cladelab(node=1256,label="2",angle=270,offset=0.2)+
  theme(legend.position = c(.05, .75))+labs(title = "NS1 Tree")

# Descendants(1255,as.phylo(NS1_tree))

NS1_node1 <- tree_subset(NS1_tree,1673,levels_back=0) %>% ggtree(aes(colour=host))+
  theme(legend.position = "none")+labs(title = "NS1 Subtree 1",subtitle = "Node 1673")+geom_rootedge(rootedge = 0.001)+border()+
  scale_colour_manual(values = c("#2CBA37","#33BFC4"),breaks = c("Dog","Horse"))
NS1_node2 <- tree_subset(NS1_tree,1256,levels_back=0) %>% ggtree(aes(colour=host))+
  theme(legend.position = "none")+labs(title = "NS1 Subtree 2",subtitle = "Node 1256")+geom_rootedge(rootedge = 0.001)+border()+
  scale_colour_manual(values = c("#F8766D","#619CFF","#F564E3"),breaks = c("Avian","Human","Pig"))

ggarrange(NS1_full,ggarrange(NS1_node1,NS1_node2,ncol=1),common.legend = T)
#ggsave("NS1 Subtrees.png",width=12,height=12,units = "cm")
```

```{r, eval=FALSE}
ns1_st1 <- ggplot(NS1_1,aes(Host,Feature,fill=mV))+geom_tile(colour="grey90")+ geom_text(aes(label = round(mV,2)),colour="black")+ scale_fill_gradient(low="white",high="forestgreen")+ labs(title="NS1 Node 1")

ns1_st2 <- ggplot(NS1_2,aes(Host,Feature,fill=mV))+geom_tile(colour="grey90")+geom_text(aes(label=round(mV,2)),colour="black")+scale_fill_gradient(low="white",high="forestgreen")+ labs(title="NS1 Node 2")

ggarrange(ns1_st1,ns1_st2,nrow=1,common.legend = T)
#ggsave("NS1 TrRates.png",width=12,height=12,units = "cm")
```

```{r PB2 subtrees, fig.height=10}
PB2_tree <- readRDS("../Phylogenies/PB2_tree.rds")
  
pb2_full <- ggtree(PB2_tree,aes(colour=host))+ 
  # geom_nodelab(aes(label=node),hjust = -0.25)+
  geom_hilight(node = 757)+geom_hilight(node = 885)+
  geom_cladelab(node=757,label="1",angle=270,offset=0.2)+
  geom_cladelab(node=885,label="2",angle=270,offset=0.2)+
  theme(legend.position = c(.05, .75))+labs(title = "PB2 Tree")

pb2_node1 <- tree_subset(PB2_tree,757,levels_back=0) %>% ggtree(aes(colour=host))+
  theme(legend.position = "none")+labs(title = "PB2 Subtree 1",subtitle = "Node 757")+geom_rootedge(rootedge = 0.001)+border()+
  scale_colour_manual(values = c("#F8766D","#F564E3"),breaks = c("Avian","Pig"))
pb2_node2 <- tree_subset(PB2_tree,885,levels_back=0) %>% ggtree(aes(colour=host))+
  theme(legend.position = "none")+labs(title = "PB2 Subtree 2",subtitle = "Node 885")+geom_rootedge(rootedge = 0.001)+border()+
  scale_colour_manual(values = c("#2CBA37","#33BFC4","#F564E3"),breaks = c("Dog","Horse","Pig"))

ggarrange(pb2_full,ggarrange(pb2_node1,pb2_node2,ncol=1),common.legend = T)
#ggsave("PB2 Subtrees.png",width=12,height=12,units = "cm")
```

```{r,eval=FALSE}
PB2_st1 <- ggplot(PB2_1,aes(Host,Feature,fill=mV))+geom_tile(colour="grey90")+ geom_text(aes(label = round(mV,2)),colour="black")+scale_fill_gradient(low="white",high="forestgreen")+ labs(title="PB2 Node 1")
PB2_st2 <- ggplot(PB2_2,aes(Host,Feature,fill=mV))+geom_tile(colour="grey90")+geom_text(aes(label = round(mV,2)),colour="black")+scale_fill_gradient(low="white",high="forestgreen")+ labs(title="PB2 Node 2")

ggarrange(PB2_st1,PB2_st2,nrow=1,common.legend = T)
#ggsave("PB2 TrRates.png",width=12,height=12,units = "cm")
```

```{r HA subtrees, fig.height=10}
HA_tree <- readRDS("../Phylogenies/HA_tree.rds")
  
HA_full <- ggtree(HA_tree,aes(colour=host))+ 
  # geom_nodelab(aes(label=node),hjust = -0.25)+
  geom_hilight(node = 3667)+geom_hilight(node = 3490)+
  geom_cladelab(node=3667,label="1",angle=270,offset=0.2)+
  geom_cladelab(node=3490,label="2",angle=270,offset=0.2)+
theme(legend.position = c(.05, .75))+labs(title = "HA Tree")

Descendants(3492,as.phylo(HA_tree))
HA_node1 <- tree_subset(HA_tree,3667,levels_back=0) %>% ggtree(aes(colour=host))+
  theme(legend.position = "none")+labs(title = "HA Subtree 1",subtitle = "Node 3667")+geom_rootedge(rootedge = 0.001)+border()+
  scale_colour_manual(values = c("#2CBA37","#33BFC4","#F564E3"),breaks = c("Dog","Horse","Pig"))
HA_node2 <- tree_subset(HA_tree,3493,levels_back=0) %>% ggtree(aes(colour=host))+geom_tiplab()+
  theme(legend.position = "none")+labs(title = "HA Subtree 2",subtitle = "Node 3493")+geom_rootedge(rootedge = 0.001)+border()+
  scale_colour_manual(values = c("#2CBA37","#619CFF","#F564E3"),breaks = c("Dog","Human","Pig"))

ggarrange(HA_full,ggarrange(HA_node1,HA_node2,ncol=1),common.legend = T)
#ggsave("HA Subtrees.png",width=12,height=12,units = "cm")
```

```{r, eval=FALSE}
HA_st1 <- ggplot(HA_1,aes(Host,Feature,fill=mV))+geom_tile(colour="grey90")+geom_text(aes(label = round(mV,2)),colour="black")+scale_fill_gradient(low="white",high="forestgreen")+ labs(title="HA Node 1")
ggsave("HA TrRates1.png",width=14,height=12,units = "cm")
HA_st2 <- ggplot(HA_2,aes(Host,Feature,fill=mV))+geom_tile(colour="grey90")+geom_text(aes(label = round(mV,2)),colour="black")+scale_fill_gradient(low="white",high="forestgreen")+ labs(title="HA Node 2")
ggsave("HA TrRates2.png",width=14,height=12,units = "cm")

ggarrange(HA_st1,HA_st2,nrow=1,common.legend = T)
#ggsave("HA TrRates.png",width=12,height=12,units = "cm")
```

```{r NA subtrees, fig.height=10}
NA_tree <- readRDS("../Phylogenies/NA_tree.rds")
  
NA_full <- ggtree(NA_tree,aes(colour=host))+ 
  # geom_nodelab(aes(label=node),hjust = -0.25)+
  geom_hilight(node = 777)+geom_hilight(node = 690)+
  geom_cladelab(node=777,label="1",angle=270,offset=0.2)+
  geom_cladelab(node=690,label="2",angle=270,offset=0.2)+
  theme(legend.position = c(.05, .75))+labs(title = "NA Tree")

# Descendants(609,as.phylo(NA_tree))

NA_node1 <- tree_subset(NA_tree,777,levels_back=0) %>% ggtree(aes(colour=host))+geom_tiplab()+
  theme(legend.position = "none")+labs(title = "NA Subtree 1", subtitle = "Node 777")+geom_rootedge(rootedge = 0.001)+
  scale_colour_manual(values = c("#F8766D","#33BFC4"),breaks = c("Avian","Horse"))+border()
NA_node2 <- tree_subset(NA_tree,690,levels_back=0) %>% ggtree(aes(colour=host))+geom_tiplab()+
  theme(legend.position = "none")+labs(title = "NA Subtree 2", subtitle = "Node 690")+geom_rootedge(rootedge = 0.001)+
  scale_colour_manual(values = c("#F8766D","#619CFF","#F564E3"),breaks = c("Avian","Human","Pig"))+border()

ggarrange(NA_full,ggarrange(NA_node1,NA_node2,ncol=1),common.legend = T)
#ggsave("NA Subtrees.png",width=12,height=12,units = "cm")
```

# Discussion

All protein sequences in the dataset were classified into one of six hosts by the array of machine learning models. These classifications, modelled exclusively from protein physio-chemical properties, were then compared with the known host from which the virus was isolated. Each value is the proportion of whole-virus sequences across the entire dataset, only groups with more than 0.01% of sequences are labelled in the heatmap. While the majority of sequences were predicted to come from the correct host (83.57% in total), the focus of this study is rather on the misclassifications; potentially indicative of signal representing species-specific adaptation, and thus taken here as a proxy for successful cross-species emergence. Hence, correct predictions are removed from the heatmap for emphasis. As an example, 2.51% of sequences ($\frac{843}{33,647}$) were predicted by the machine learning models to be swine IAV when they were in fact sampled from avian hosts. Similarly, 8.31% ($\frac{2737}{33,647}$) of sequences were mislabelled as human-origin viruses despite originating from swine.

The phylogenetic modelling rather calculates the transition rate of sequences between labelled metadata groups (i.e. host). Given the phylogenetic structure of the samples, and the labels assigned to each tip (the origin host), the likelihood of sequences moving between classes (i.e. sustained cross-species emergence and adaptation) is given as an expected transition rate. For example, the rate of bat virus emergence in avian populations is modelled as very unlikely (0.19), whereas viruses from pigs appear highly liable to spillover **and then adapt** into human hosts (18.2). Both of these predictions align with the observed ecology of IAV over the past century and a half; swine viruses have repeatedly crossed into human populations. To note, these transition rates are not estimated to be symmetrical – the flow of IAV between humans and swine is heavily unidirectional, with a rate of 3.76 in human → swine emergence compared to the above example (18.2 in swine → human cross-species transmission).

In the future, we would’ve clustered per host and also possibly split the birds up more into ‘waterfowl vs others’ instead of assuming that all birds are always the source of infection. Other potential subdivisions include the splitting of human H1N1 viruses into two distinct classes of pre- and post-2009 samples in order to distinguish already circulating viruses from those that emerged during the 2009 swine flu pandemic.

## Sequence Reconstruction

In order to test the models further, protein features of the internal nodes could be calculated. Of course, the majority of these internal nodes do not have sequences collected. To circumvent this, a further step of our analyses could be to reconstruct ancestral sequences. By estimating the sequence of an internal node, it can then be re-input into iFeaturesOmega. Then, if our hypothesis stands, the features of these inferred proteins ought to be an average of the two descendants. CodeML is suggested as a tool for reconstructing ancestral protein sequences.


# References

<div id="refs"></div>

# Appendix

```{r}
Clustering_Tests <- read_xlsx("../Sequences/Clustering Tests.xlsx",sheet = "Clustering Tests")
Clustering_Tests$Identity <- Clustering_Tests$Identity %>% as.factor()
totals <- Clustering_Tests %>% subset(Identity==100)
totals$CoV <- "cov0"
Clustering_Tests2 <- rbind(Clustering_Tests,totals) %>%
  pivot_wider(names_from = c("Coverage","Identity"), values_from = Count) %>% 
  select(order(colnames(.))) %>% relocate("Group") %>% relocate("Protein", .after="Group") %>% relocate("CoV", .after="Protein") %>% relocate("c80_100", .after = last_col())

Clustering_Tests2 %>% subset(Group=="Avian", select=-1) %>% 
  kbl(align='c',caption="Avian",col.names = c("Protein","Mode","75","85","90","95","99","75","85","90","95","99","75","85","90","95","99","Original")) %>% kable_styling() %>% collapse_rows(c(1,18)) %>% add_header_above(c(" "=1," "=1,"c50"=5,"c70"=5,"c80"=5," "=1))
```


```{r}
Clustering_Tests2 %>% subset(Group=="Mammal", select=-1) %>%
  kbl(align='c',caption="Mammal",col.names = c("Protein","Mode","75","85","90","95","99","75","85","90","95","99","75","85","90","95","99","Original")) %>% kable_styling() %>% collapse_rows(c(1,18)) %>% add_header_above(c(" "=1," "=1,"c50"=5,"c70"=5,"c80"=5," "=1))
```

